A game-theory modeling approach to utility and strength of interactions dynamics in biomedical research social networks

### 2017-04-03 casm.tex
fixed som reference, created research areas table.

\begin{abstract} % abstract
%\parttitle{First part title} %if any
Collaboration has become a cornerstone in biomedical research today.
In contrast to physics which has a long history and experience in
collaborative projects, biology is only recently becoming an evermore
collaborative discipline.  In this article we explore the effect of a
collaboration network on the distribution of players having access to
certain amount of resources from other players in the network and the
distribution of the strength of interactions among them. Particularly,
we implemented two games played simultaneously: one for maximizing
individual utility based on the iterated Prisoner's Dilemma; the
other, a coordination game for maximizing the connection strength
between players. We are interested in how they affect each other in
the context of a network of scientific collaboration under the idea
that while researchers are interested in maximizing their utilities,
they also know that it is important to invest in building
collaborative relationships.  We tested our simulation on a biomedical
research community network in M\'exico and compared the results with
an Erd\"{o}s-Reny\'i, a Watts-Strogatz small-world and
Barab\'asi-Albert topologies. Different topologies display different
global utility and global strength of interaction
distributions. Moreover, the distribution of utility and strength of
interaction in the researchers network is similar to that of
Barab\'asi-Albert and Watts-Strogatz topologies, respectively. We
believe that utility distribution in the researchers network suggests
that there are socio-cultural mechanisms governing the network that
produce an asymmetric distribution of resources. The high distribution
of strong connections might reflect some sort of subordination among
researchers by which they are morally obliged to cooperate by the same
socio-cultural mechanisms. Finally, the range around the threshold
that regulates the decision to cooperate or defect according to the
agent's historical balance between utility and strength of
collaborative relationships and carrying capacity of the system is
small, suggesting that there is a region in which a phase transition
takes place from a population of cooperators to a population of
defectors. Simulations like ours may help to develop science policies
to promote fair distribution of resources.

@@

Even though curation and analysis work of this database is still going
on, some relevant facts about the biomedical research can be
said. Over the period of 12 years, 32 general research areas have been
defined, the three most funded research areas are \emph{chronic and
  degenerative diseases}, \emph{malignant neoplasms}, and
\emph{infectious and parasitic diseases}. The least funded area is
\emph{Ethics and medicine}. The area with the most researchers is
\emph{malignant neoplasms}. Other areas of relevance for M\'exico are
\emph{diseases related to poverty} and \emph{Health and vulnerable
  groups}. \\

@@

The probability for an agent to cooperate or defect depends on a number ($\eta$)
that refers to a historical balance between average utility and the average
strength of the connections with its neighbors. It is modeled so because we
assume that whatever the result in utility or strength of interaction, as long
as one of them increases, the player will be confident in the strategy followed
so far. {\color{red}In every simulation step, each agent will independently play both games
with every neighbor. Nonetheless, the agent sets its decision to cooperate or
defect beforehand by assessing the overall situation of its relationships. This
may sound odd, since one may think that a person decides for each relationship
separately. However in this way players are optimizing access to resources and
the strength of interactions to their neighbors at the same time, since their
state is set by considering both together. The strength of interaction is
updated independently for each pair of agents.  In a single play some
interactions might be weakened, some may get strengthened and others may remain
unaffected. Moreover, in this model, history is kept in the network as strength
of interactions and it conditions the agent\'s disposition to cooperate or to
defect. As the sayings go: Burnt children dread fire; A bad dog is not born but made.}\\

@@

For the agent to decide whether to cooperate or not, $\eta$ is
compared to a global threshold $\nu$. If the agent's $\eta$ $>$ $\nu$,
then he will cooperate, otherwise he will be suspicious and
will defect. $\nu$ is a global parameter that establishes a threshold
that an agent's $\eta$ must cross in order to decide to cooperate. In
this way, $\nu$ can limit the size of the population of
cooperators. Due to what the system and the game can offer to agents
in terms of utilities and the strength of collaboration relationships,
$\nu$ represents the carrying capacity of the
system for the population of cooperators.\\

@@

We ran two different experiments. In the First Experiment we simulated different
values of carrying capacity $\nu$. With this experiment we were able
to see how the number of cooperators, utility, strength of interactions
among agents and the ratio of shifting state population would change
in the range of the carrying capacity. The states of the agents were
the same at initialization, for all values of the carrying
capacity. Since the model is deterministic, it will return the same
result if run under the same conditions.\\

The Second Experiment consisted in running the simulation under the
same degree of carrying capacity $\nu$ but randomizing the initial
states of the agents. This would show that the system converges to a
global state. For every network, the simulation was run 100 times and
results were averaged.

@@

Other results to report are those of the dynamics of different
variables as the carrying capacity $\nu$ changes. The  most salient
result is that for $\nu$ between $0.19$ and $0.24$, there is an
apparent phase transition in all different topologies and for all the
different variables. It is worth noting that the shape
of the phase transition-like behavior is different according to the topology of the
network under study. When $\nu$ is between $0.0$ and $0.2$, that is, when there is
no space or a very short one for suspiciousness, all agents cooperate, when
$\nu$ is above $0.25$, all agents defect. \textit{Utility},
\textit{strength of interactions}, and \textit{changing state population}
replicate that same behavior for the same limits. \\

@@

\begin{figure} [h!]
\centering
\begin{tabular}{cc}
%   \includegraphics[scale=0.28]{images/erdos_dc_nus_1.png} & \includegraphics[scale=0.28]{images/watts_dc_nus_1.png}
%   \\
%   \includegraphics[scale=0.28]{images/barabasi_dc_nus_1.png} & \includegraphics[scale=0.28]{images/fosiss_dc_nus_1.png}
\end{tabular}
\caption{Ratio of cooperators as a function of carrying capacity and time during First Experiment. Topologies are: A. Random, Erd\"{o}s-R\'enyi. B. Watts-Strogatz. C. Barab\'asi-Albert. D. FOSISS.
}\label{CD}
\end{figure}

@@

% \includegraphics[scale=0.28]{images/erdos_fitness_nus_1.png} & \includegraphics[scale=0.28]{images/watts_fitness_nus_1.png} \\
% \includegraphics[scale=0.28]{images/barabasi_fitness_nus_1.png} & \includegraphics[scale=0.28]{images/fosiss_fitness_nus_1.png}
\end{tabular}
\caption{First Experiment's utility $U$ dynamics as a function of carrying capacity $\nu$ and time. Topologies are: A. Random, Erd\"{o}s-R\'enyi. B. Watts-Strogatz. C. Barab\'asi-Albert. D. FOSISS.}\label{fitness}
\end{figure}

@@

% \includegraphics[scale=0.28]{images/erdos_trust_nus_1.png} & \includegraphics[scale=0.28]{images/watts_trust_nus_1.png} \\
% \includegraphics[scale=0.28]{images/barabasi_trust_nus_1.png} & \includegraphics[scale=0.28]{images/fosiss_trust_nus_1.png}
\end{tabular}
\caption{Strength of interactions $w$ dynamics as a function of carrying capacity $\nu$ and time during First Experiment. Topologies are: A. Random, Erd\"{o}s-R\'enyi. B. Watts-Strogatz. C. Barab\'asi-Albert. D. FOSISS.}\label{trust}
\end{figure}

@@

% \includegraphics[scale=0.28]{images/erdos_state_nus.png} & \includegraphics[scale=0.28]{images/watts_state_nus.png} \\
% \includegraphics[scale=0.28]{images/barabasi_state_nus.png} & \includegraphics[scale=0.28]{images/fosiss_state_nus.png}
\end{tabular}
\caption{Shifting population between cooperators and defectors as a function of
carrying capacity $\nu$ and time during First Experiment. Topologies are: A. Random, Erd\"{o}s-R\'enyi. B. Watts-Strogatz. C. Barab\'asi-Albert. D. FOSISS.}\label{state} 
\end{figure}

@@

 To close this article, we would like to comment on three
  issues we consider important. The first one is about confronting our
  results with data available to us from the biomedical research
  community. Second, some remarks we believe are important about the
  role of computational simulations in the social sciences. Finally we
  would like to comment on our future work.\\

  The two main results of our model are the smoothness in
  the phase transition-like behavior for different parameters and the
  distribution of utility and strength of interactions in the FOSISS
  network compared to other networks and their topologies. We are
  certain that the particular structure of FOSISS network is playing a
  central role in the results and because of that we would like to
  discuss it a little bit further.\\

  FOSISS network is highly hierarchical according to its
  heterogeneity of $0.873$. Because of it, one would expect to find
  the presence of important hubs \cite{Wu:2008}, that is, a few
  researchers control the whole network, as has been reported in some
  other places \cite{Yousefi-etal:2008}. Surprisingly, FOSISS network
  centralization is very low $0.023$, i.e. there are no researchers
  that centralize the majority of connections. Our guess is that the
  network is composed of many small communities or groups with a
  central researcher or Principal Investigator (PI). If this is the
  case in the FOSISS network, it means that those groups have a very
  hierarchical structure as well.\\

@@

 As is common among scientific communities in biomedical
  research, PIs play a central role in the network. Strong PIs and
  well connected groups seem to be somehow responsible for the high
  levels of strength of interactions and the centralization of utility
  in our simulation. As for what seems to be phase transitions, in the
  case of FOSISS networks, these are smoother than those in the other
  networks with different topologies, even for those with a
  small-world topology. We think that this behavior is also the result
  of the hierarchical structure already mentioned. If this is true,
  strength of interactions is first lost in the edges that link
  different groups and then in the edges that connect members of the
  groups. Connections between groups would not be as dense as those
  inside the groups, which means that there would not be enough
  information on the behavior of one group regarding its neighbors to
  constrain them as it seems to happen with individual researchers
  inside their communities. Nevertheless, if values of the carrying
  capacity $\nu$ keep increasing and it becomes more difficult to
  strengthen interactions, then strength of interactions begins to
  diminish inside groups.\\

>>>>

\begin{abstract} % abstract
%\parttitle{First part title} %if any
Collaboration has become a cornerstone in biomedical research today.
In contrast to physics which has a long history and experience in
collaborative projects, biology is only recently becoming an evermore
collaborative discipline.  In this article we explore the effect of a
collaboration network on the distribution of players having access to
certain amount of resources from other players in the network and the
distribution of the strength of interactions among them. Particularly,
we implemented two games played simultaneously: one for maximizing
individual utility based on the iterated Prisoner's Dilemma; the
other, a coordination game for maximizing the connection strength
between players. We are interested in how they affect each other in
the context of a network of scientific collaboration under the idea
that while researchers are interested in maximizing their utilities,
they also know that it is important to invest in building
collaborative relationships.  We tested our simulation on a biomedical
research community network in M\'exico and compared the results with
an Erd\"{o}s-Reny\'i, a Watts-Strogatz small-world and
Barab\'asi-Albert topologies. Different topologies display different
global utility and global strength of interaction
distributions. Moreover, the distribution of utility and strength of
interaction in the researchers network is similar to that of
Barab\'asi-Albert and Watts-Strogatz topologies, respectively. We
believe that utility distribution in the researchers network suggests
that there are socio-cultural mechanisms governing the network that
produce an asymmetric distribution of resources. The high distribution
of strong connections might reflect some sort of subordination among
researchers by which they are morally obliged to cooperate by the same
socio-cultural mechanisms. Finally, the range around the threshold
that regulates the decision to cooperate or defect according to the
agent's historical balance between utility and strength of
collaborative relationships and an a parameter of confidence of the system is
small, suggesting that there is a region in which a phase transition
takes place from a population of cooperators to a population of
defectors. Simulations like ours may help to develop science policies
to promote fair distribution of resources.

@@

Even though curation and analysis work of this database is still going
on, some relevant facts about the biomedical research can be
said. Over the period of 12 years, 37 general research areas have been
defined, the three most funded research areas are \emph{chronic and
  degenerative diseases}, \emph{malignant neoplasms}, and
\emph{infectious and parasitic diseases}. The least funded area is
\emph{Ethics and medicine}. The area with the most researchers is
\emph{malignant neoplasms}. Other areas of relevance for M\'exico are
\emph{diseases related to poverty} and \emph{Health and vulnerable
  groups}. \\

@@

The probability for an agent to cooperate or defect depends on a number ($\eta$)
that refers to a historical balance between average utility $\langle u_i \rangle$ and the average
strength of the connections with its neighbors $\langle w_{ij} \rangle _j$. It is modeled so because we
assume that whatever the result in utility or strength of interaction, as long
as one of them increases, the player will be confident in the strategy followed
so far. {\color{red}In every simulation step, each agent will independently play both games
with every neighbor. Nonetheless, the agent sets its decision to cooperate or
defect beforehand by assessing the overall situation of its relationships. This
may sound odd, since one may think that a person decides for each relationship
separately. However in this way players are optimizing access to resources and
the strength of interactions to their neighbors at the same time, since their
state is set by considering both together. The strength of interaction is
updated independently for each pair of agents.  In a single play some
interactions might be weakened, some may get strengthened and others may remain
unaffected. Moreover, in this model, history is kept in the network as strength
of interactions and it conditions the agent\'s disposition to cooperate or to
defect. As the sayings go: Burnt children dread fire; A bad dog is not born but made.}\\

@@

For the agent to decide whether to cooperate or not, $\eta$ is compared to a
global parameter $\nu$. This is a parameter that represents the minimum value of
$\eta$ that a player must achieve in order to cooperate. If the agent's
$\eta$ $>$ $\nu$, then he will cooperate, otherwise he will be suspicious and
will defect. In this way, as we arbitrarily control the value of $\nu$ we can
identify the values of $\eta$ where the population stops cooperating and begins
to defect; also it makes possible to see how this transition takes place for the
different network topologies.  \\

%{\color {red}Due to what the system and the game can offer to agents
%in terms of utilities and the strength of collaboration relationships,
%$\nu$ represents a parameter of confidence of the system.}\\

@@

We ran two different experiments. In the First Experiment we simulated different
values of the parameter of confidence $\nu$. With this experiment we were able
to see how the number of cooperators, utility, strength of interactions
among agents and the ratio of shifting state population would change
in the range of the parameter of confidence. The states of the agents were
the same at initialization, for all values of $\nu$. Since the model is deterministic, it will return the same
result if run under the same conditions.\\

The Second Experiment consisted in running the simulation under the
same value of the parameter of confidence $\nu$ but randomizing the initial
states of the agents. This would show that the system converges to a
global state. For every network, the simulation was run 100 times and
results were averaged.

@@

Other results to report are those of the dynamics of different
variables as the parameter of confidence $\nu$ changes. The  most salient
result is that for $\nu$ between $0.19$ and $0.24$, there is an
apparent phase transition in all different topologies and for all the
different variables. It is worth noting that the shape
of the phase transition-like behavior is different according to the topology of the
network under study. When $\nu$ is between $0.0$ and $0.2$, that is, when there is
no space or a very short one for suspiciousness, all agents cooperate, when
$\nu$ is above $0.25$, all agents defect. \textit{Utility},
\textit{strength of interactions}, and \textit{changing state population}
replicate that same behavior for the same limits. \\

@@

\begin{figure} [h!]
\centering
\begin{tabular}{cc}
%   \includegraphics[scale=0.28]{images/erdos_dc_nus_1.png} & \includegraphics[scale=0.28]{images/watts_dc_nus_1.png}
%   \\
%   \includegraphics[scale=0.28]{images/barabasi_dc_nus_1.png} & \includegraphics[scale=0.28]{images/fosiss_dc_nus_1.png}
\end{tabular}
\caption{Ratio of cooperators as a function of the parameter of confidence and time during First Experiment. Topologies are: A. Random, Erd\"{o}s-R\'enyi. B. Watts-Strogatz. C. Barab\'asi-Albert. D. FOSISS.
}\label{CD}
\end{figure}

@@

% \includegraphics[scale=0.28]{images/erdos_fitness_nus_1.png} & \includegraphics[scale=0.28]{images/watts_fitness_nus_1.png} \\
% \includegraphics[scale=0.28]{images/barabasi_fitness_nus_1.png} & \includegraphics[scale=0.28]{images/fosiss_fitness_nus_1.png}
\end{tabular}
\caption{First Experiment's utility $U$ dynamics as a function of the parameter
  of confidence $\nu$ and time. Topologies are: A. Random, Erd\"{o}s-R\'enyi. B. Watts-Strogatz. C. Barab\'asi-Albert. D. FOSISS.}\label{fitness}
\end{figure}

@@

% \includegraphics[scale=0.28]{images/erdos_trust_nus_1.png} & \includegraphics[scale=0.28]{images/watts_trust_nus_1.png} \\
% \includegraphics[scale=0.28]{images/barabasi_trust_nus_1.png} & \includegraphics[scale=0.28]{images/fosiss_trust_nus_1.png}
\end{tabular}
\caption{Strength of interactions $w$ dynamics as a function of the parameter
  of confidence $\nu$ and time during First Experiment. Topologies are:
  A. Random, Erd\"{o}s-R\'enyi. B. Watts-Strogatz. C. Barab\'asi-Albert. D. FOSISS.}\label{trust}  
\end{figure}

@@

% \includegraphics[scale=0.28]{images/erdos_state_nus.png} & \includegraphics[scale=0.28]{images/watts_state_nus.png} \\
% \includegraphics[scale=0.28]{images/barabasi_state_nus.png} & \includegraphics[scale=0.28]{images/fosiss_state_nus.png}
\end{tabular}
\caption{Shifting population between cooperators and defectors as a function of
the parameter
  of confidence $\nu$ and time during First Experiment. Topologies are:
  A. Random,
  Erd\"{o}s-R\'enyi. B. Watts-Strogatz. C. Barab\'asi-Albert. D. FOSISS.}\label{state}  
\end{figure}

@@

  There are two issues we consider important. The first one is about
  confronting our results with data available to us from the biomedical research
  community. Second, we would like to comment on our future work.\\

  The two main results of our model are the smoothness in the phase
  transition-like behavior for different parameters and the distribution of
  utility and strength of interactions in the FOSISS network compared to other
  networks and their topologies. We are certain that the particular structure of
  FOSISS network is playing a central role in the results and because of that we
  would like to discuss it a little bit further. FOSISS network is highly
  hierarchical according to its heterogeneity of $0.873$. Because of it, one
  would expect to find the presence of important hubs \cite{Wu:2008}, that is, a
  few researchers control the whole network, as has been reported in some other
  places \cite{Yousefi-etal:2008}. Surprisingly, FOSISS network centralization
  is very low $0.023$, i.e. there are no researchers that centralize the
  majority of connections. Our guess is that the network is composed of many
  small communities or groups with a central researcher or Principal
  Investigator (PI). If this is the case in the FOSISS network, it means that
  those groups have a very hierarchical structure as well.\\

@@

 As is common among scientific communities in biomedical
  research, PIs play a central role in the network. Strong PIs and
  well connected groups seem to be somehow responsible for the high
  levels of strength of interactions and the centralization of utility
  in our simulation. As for what seems to be phase transitions, in the
  case of FOSISS networks, these are smoother than those in the other
  networks with different topologies, even for those with a
  small-world topology. We think that this behavior is also the result
  of the hierarchical structure already mentioned. If this is true,
  strength of interactions is first lost in the edges that link
  different groups and then in the edges that connect members of the
  groups. Connections between groups would not be as dense as those
  inside the groups, which means that there would not be enough
  information on the behavior of one group regarding its neighbors to
  constrain them as it seems to happen with individual researchers
  inside their communities. Nevertheless, if values of the parameter
  of cooperation $\nu$ keep increasing and it becomes more difficult to
  strengthen interactions, then strength of interactions begins to
  diminish inside groups.\\
